Abstractions for Multi-Sorted Substitutions

### 2023-09-18 paper/paper.tex
Update.

  Additionally, repetition can also occur within a single formalization.
  This can be seen with polymorphic languages, where multiple sorts
  of variables are present.
  Consider for example System F, which supports both expression- and
  type-variables.
  With a naive approach, the whole substitution machinery needs to be
  duplicated three times! We need to substitute expression-variables in
  expressions, type-variables in types, but we also need to substitute
  type-variables in expressions.
  Even worse, having two substitutions acting on expressions, requires
  to also prove lemmas about their interactions.
  If we would additionally introduce kind-variables, we would end up
  with a total of six duplications of the substition machinery and
  corresponding interaction lemmas!

@@

  Even more repetition is introduced by having a typed language,
  where for each substitution (and renaming) operation, we need to prove
  that the operation preserves typing, again doubling the repetition.

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  In summary, formalizing a language with expression-, type-, and
  kind-variables with a naive approach can end up with a total of 24
  repetitions for each definition related to substitution, not counting
  the interaction lemmas between different substitutions that are
  acting on the same syntactic categories!

@@

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      Appendix TODO explains the workaround.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

@@

  We then define a \ACode{\ACon{TypingTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTypingTraversal
  Given a variable/term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and then later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases they are formulated for intrinsic typing.

@@

  \section{Future Work}
  \label{sec:future}
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Incorporating a decision procedure for equational
    substitution-lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}

@@

>>>>

  Additionally, repetition can also occur within a single formalization.
  This can be seen with polymorphic languages, where multiple sorts
  of variables are present.
  Consider for example System F, which supports both expression- and
  type-variables.
  With a naive approach, the whole substitution machinery needs to be
  duplicated three times! We need to substitute expression-variables in
  expressions, type-variables in types, but also type-variables in
  expressions.
  Even worse, having two substitutions acting on expressions, requires
  to also prove lemmas about their interactions.
  If we would additionally introduce kind-variables, we would end up
  with a total of six duplications of the substition machinery and
  corresponding interaction lemmas!

@@

  Similarly, for a typed language formalized via extrinsic typing we
  need type preservation lemmas for each substitution and renaming
  operation, which again doubles the repetition.

  % In summary, to have enough substitution machinery to prove subject
  % reduction for a language with expression-, type-, and
  % kind-variables one can end up with a total of 24
  % repetitions for each definition related to substitution, not counting
  % the interaction lemmas between different substitutions that are
  % acting on the same syntactic categories!
  If one is not careful, a formalization with three sorts of variables
  can easily end up with 24 definitions for each operation related
  to substitution, which usually includes at least a substitution type, and
  operations for application, extension, weakening, lifting, singleton
  substitution, extensional equality, and an identity substitution.

  In summary, formalizing a language with expression-, type-, and
  kind-variables with a naive approach can end up with a total of 24
  repetitions for each definition related to substitution, not counting
  the interaction lemmas between different substitutions that are
  acting on the same syntactic categories!

@@

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

@@

  We then define a \ACode{\ACon{TypingTraversal}} analogously to
  \ACode{\ACon{Traversal}}, but instead of defining the application of
  maps, it defines the type preservation of applying maps:
  \LibCode\KTypingTraversal
  Given a term \ACode{e} with typing \ACode{⊢e} and
  a renaming/substitution \ACode{ϕ} with typing \ACode{⊢ϕ},
  the term \ACode{⊢e \AField{⊢⋯} ⊢ϕ} is a typing for \ACode{e \AField{⋯} ϕ}.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases they are formulated for intrinsic typing.

@@

  \section{Future Work}
  \label{sec:future}
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Extending the class of typing relations covered by
    the \ACode{\ACon{TypingTraversal}} abstraction and identifying
    a subclass for which it can be derived.
  \item
    Incorporating a decision procedure for equational
    substitution-lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}

@@

  \section{System F with Reflection}
  \label{sec:reflection-example}

  \section{Map Equality Without Functional Extensionality}
  \label{sec:fun-ext}

  \section{Features of the Actual Implementation}
  \label{sec:features}
